Supplememtary examples for:

### 2020-07-09 extensions.Rmd
Made minor edits to the text

---
title: "Supplememtary examples for:"
#author: "Henrik Kenneth Andersen"
#date: "`r Sys.Date()`"
output: 
  pdf_document:
    fig_caption: yes
    number_sections: yes
    includes: 
      in_header: header.tex
documentclass: article
classoption: a4paper
bibliography: references.bib   
---

@@

The following discusses several opportunities to extend the basic FE model outlined in the main article, and provides concrete guidance on the implementation in `lavaan`. First, I go over a number of possibilities to relax assumptions associated with the traditional FE model. Then, I discuss the issue of measurement error and show how we can use latent variables to deal with it and allow us to properly estimate the coefficients of interest. Then, I show display a type of hybrid FE/RE model that allows us to control for time-invariant unobserved heterogeneity that allows us to include time-invariant predictors in the model.

@@

There are a number of implicit assumptions attached to the typical FE model that can be relaxed in SEM. Some of these assumptions have been discussed already, and a full list of assumptions can be found in @Bollen2010. Here, I will go over just a few, concentrating on the implementation in `lavaan` and the opportunity to empirically test whether the adjustments are justified or not.

@@

For example, we can rewrite the original FE equation as
\begin{align}
y_{it} & = \beta_{t}x_{it} + \lambda_{t}\alpha_{i} + \varepsilon_{it}
\end{align}
where $\beta$ becomes $\beta_{t}$ and the implicit regression weight of one turns to $\lambda_{t}$ to highlight the fact that the effect of $x$ as well as $\alpha$ on $y$ may vary over time. We can furthermore easily relax the assumption of time-constant error variance, i.e., $\sigma^{2}_{\varepsilon_{t}}$. As noted above, the assumption regarding $\E[\alpha x_{t}]$ in $\bm{\Psi}$ determines whether we have an FE or RE model. We can set these to zero and test whether the RE model would be preferable to the FE model. In general, if the individual effects are truly uncorrelated with the model covariates, it is advisable to switch to an RE model since because it uses up less degrees of freedom, it will have smaller asymptotic standard errors [@Bollen2010].

@@

As outlined in the @Bollen2010 text, the researcher has the opportunity to test each of the assumptions empirically and decide whether a more parsimonious, i.e., restrictive model is justifiable. For each assumption, a likelihood ratio test can be carried out to determine whether the improvement to model fit resulting from the relaxation of various assumptions is significant or whether the more parsimonious model is preferable after all.

If we use the original model `fe_sem.fit` as a starting point, the best strategy for testing these assumptions is to work in a stepwise fashion, relaxing one assumption at a time. We can begin by first constraining the correlation between $\alpha$ and $x_{t}$ to zero (`re_sem`) for an RE model. If turning from an FE to an RE model does not significantly worsen model fit, we can go forward with the rest of the steps with the RE model. If, however, the fit does worsen significantly, it is likely better to stick with the FE model; moving forward then with it to see if a less restrictive FE model is preferable. We can perform a likelihood ratio test in `R` using the `anova()` function:

@@

To demonstrate the effect of measurement error on the FE-SEM model, and then provide a strategy for dealing with measurement error in SEM, let us first modify the simulated dataset by generating multiple *indicators* of the independent and dependent variables that all measure the intended variable imprecisely. Returning to our panel data, we have three indicators of each the independent and dependent variable, per timepoint:  
\begin{align}
\tilde{x}_{kt} & = x_{t} + \upsilon_{kt}, \\
\tilde{y}_{kt} & = y_{t} + \nu_{kt}
\end{align}
where $k = 1, 2, 3$ and $t = 1, ..., T$. This is like repeatedly presenting a respondent with a multi-item scale designed to measure things like stress, depression, xenophobia, etc. over the course of a panel study. To create the observed indicators, a random amount of measurement error (ranging from $\{\sigma^{2}_{\upsilon_{k}}, \sigma^{2}_{\nu_{k}}\}\in \{1.0, 1.1, 1.2, 1.3, 1.4, 1.5\}$) was added to the true variables. The code for generating impercise indicators can be found in Appendix B.

Let us first focus on the issue of imprecise measurements of the independent variable of interest and run the same FE-SEM model above, but this time we will use one of the measurement error sullied indicators, here $\tilde{x}_{1t}$, instead of the true independent variable, $x_{t}$.

@@

Thus, our FE regression equation changes from $y_{t} = \beta x_{t} + \alpha + \varepsilon_{t}$ to:
\begin{align}
\eta_{t} & = \beta \xi_{t} + \alpha + \zeta_{t}
\end{align}
where $\zeta_{t}$ represents the disturbance, in other words the residual of the latent dependent variable $\eta_{t}$.

First, however, let us double-check that measurement error in the dependent variable only increases the error variance (thus also increasing standard errors and reducing $R^{2}$), but does not systematically bias the coefficients of interest. The next model uses the indicators of $x$ and specifies latent variables to represent the valid cross-sectional variance. The dependent variable in the model is one of the impercisely measured indicators of $y$.

@@

Finally, to see the benefits of removing measurement error from the dependent variable in terms of standard errors and $R^{2}$ statistics, we can specify a model with latent variables representing the valid cross-sectional variance in $y$.

@@

Generally, the use of a latent variable to account for measurement error in the dependent variable will help smooth out chance variation in the estimated coefficient and increase model efficiency, i.e., make the estimated coefficient closer to the true parameter more often. The standard errors will tend to be smaller compared to a model in which the dependent variable still contains measurement error (e.g., `fe_sem4.fit`: `r round( lavInspect( fe_sem4.fit, "list")[ 36, 15], 3)` vs. `fe_sem3.fit`: `r round( lavInspect( fe_sem3.fit, "list")[ 21, 15], 3)`) and the model fit will be better.

@@

It is important, however, to realize that the unbiasedness of $\hat{\gamma}$ in this model is dependent on the assumption that $\E[\zeta | \bm{\xi_{t}}, \alpha_{2}] = 0$. In other words, the idiosyncratic error is mean independent of $\bm{\xi_{t}} = \xi_{1}, \xi_{2}, ..., \xi_{T}$ as well as $\alpha_{2}$. The first part is easier to accept because we are controlling for all potential time-invariant confounders that could induce a relationship between the independent variable and the error. The unbiasedness of $\hat{\gamma}$, on the other hand rests on the assumption that the time-invariant predictor is independent of the error. If $\alpha_{2}$ represented the respondent's intelligence and $\eta_{t}$ represented the respondent's income, for example, then $\hat{\gamma}$ would be biased if both were dependent on a third time-invariant variable, say level of schooling. For this reason, we need to treat the regression on a time-invariant predictor like any other regular multivariate regression model and look to include all plausible potential confounders as controls in the model, or turn to other methods, e.g., instrumental variables.

>>>>

---
title: "Supplememtary examples for:"
#author: "Henrik Kenneth Andersen"
#date: "`r Sys.Date()`"
output: 
  pdf_document:
    fig_caption: yes
    number_sections: yes
    includes: 
      in_header: header.tex
documentclass: article
classoption: a4paper
urlcolor: blue
link-citations: yes
bibliography: references.bib   
---

@@

The following discusses several opportunities to extend the basic FE model outlined in the main article, and provides concrete guidance on the implementation in `lavaan`. First, I go over a number of possibilities to relax assumptions associated with the traditional FE model. Then, I discuss the issue of measurement error and show how we can use latent variables to deal with it and properly estimate the coefficients of interest. Then, I show a type of hybrid FE/RE model that allows us to control for time-invariant unobserved heterogeneity while including time-invariant predictors in the model.

@@

There are a number of implicit assumptions attached to the typical FE model that can be relaxed in SEM. Some of these assumptions have been discussed already, and a fairly comprehenisve list of assumptions can be found in @Bollen2010. Here, I will go over just a few, concentrating on the implementation in `lavaan` and the opportunity to empirically test whether the adjustments are justified or not.

@@

For example, we can rewrite the original FE equation as
\begin{align}
y_{it} & = \beta_{t}x_{it} + \lambda_{t}\alpha_{i} + \varepsilon_{it}
\end{align}
where $\beta$ becomes $\beta_{t}$ and the implicit regression weight of one turns to $\lambda_{t}$ to highlight the fact that the effect of $x$ as well as $\alpha$ on $y$ may vary over time. We can furthermore easily relax the assumption of time-constant error variance, i.e., $\sigma^{2}_{\varepsilon_{t}}$. As noted in the [main article](https://github.com/henrik-andersen/FE-SEM/blob/master/article.pdf), the assumption regarding $\E[\alpha x_{t}]$ in $\bm{\Psi}$ determines whether we have an FE or RE model. We can set these to zero and test whether the RE model would be preferable to the FE model. In general, if the individual effects are truly uncorrelated with the model covariates, it is advisable to switch to an RE model since because it uses up less degrees of freedom, it will have smaller standard errors [@Bollen2010].

@@

As outlined in @Bollen2010, the researcher has the opportunity to test each of the assumptions empirically and decide whether a more parsimonious, i.e., restrictive model is justifiable. For each assumption, a likelihood ratio test can be carried out to determine whether the improvement to model fit resulting from the relaxation of various assumptions is significant or whether the more parsimonious model is preferable after all.

If we use the original model `fe_sem.fit` (from the [main article](https://github.com/henrik-andersen/FE-SEM/blob/master/article.pdf)) as a starting point, the best strategy for testing these assumptions is to work in a stepwise fashion, relaxing one assumption at a time. We can begin by first constraining the correlation between $\alpha$ and $x_{t}$ to zero (`re_sem`) for an RE model. If turning from an FE to an RE model does not significantly worsen model fit, we can go forward with the rest of the steps with the RE model. If, however, the fit does worsen significantly, it is likely better to stick with the FE model; moving forward then with it to see if a less restrictive FE model is preferable. We can perform a likelihood ratio test in `R` using the `anova()` function:

@@

To demonstrate the effect of measurement error on the FE-SEM model, and then provide a strategy for dealing with measurement error in SEM, the [simulated dataset](https://github.com/henrik-andersen/FE-SEM/blob/master/simulation-code.R) generates multiple *indicators* of the independent and dependent variables that all measure the intended variable imprecisely. Returning to our panel data, we have three indicators of each the independent and dependent variable, per timepoint:  
\begin{align}
\tilde{x}_{kt} & = x_{t} + \upsilon_{kt}, \\
\tilde{y}_{kt} & = y_{t} + \nu_{kt}
\end{align}
where $k = 1, 2, 3$ and $t = 1, ..., T$. This is like repeatedly presenting a respondent with a multi-item scale designed to measure things like stress, depression, xenophobia, etc. over the course of a panel study. To create the observed indicators, a random amount of measurement error (ranging from $\{\sigma^{2}_{\upsilon_{k}}, \sigma^{2}_{\nu_{k}}\}\in \{1.0, 1.1, 1.2, 1.3, 1.4, 1.5\}$) was added to the true variables, again see [the simulation code](https://github.com/henrik-andersen/FE-SEM/blob/master/simulation-code.R).

Let us first focus on the issue of imprecise measurements of the independent variable of interest and run the same FE-SEM model above, but this time we will use one of the measurement error sullied indicators, here $\tilde{x}_{1t}$, instead of the true independent variable, $x_{t}$. As for the naming conventions in the `R` code, `x11` stands for the first indicator ($k = 1$) at the first point in time ($t = 1$), whereas for example `x35` stands for the third indicator ($k = 3$) at the fifth point in time ($t = 5$).

@@

Thus, our FE regression equation changes from $y_{t} = \beta x_{t} + \alpha + \varepsilon_{t}$ to:
\begin{align}
\eta_{t} & = \beta \xi_{t} + \alpha + \zeta_{t}
\end{align}
where $\zeta_{t}$ represents the disturbance, in other words the residual of the latent dependent variable $\eta_{t}$. 
First, however, let us double-check that measurement error in the dependent variable only increases the error variance (thus also increasing standard errors and reducing $R^{2}$), but does not systematically bias the coefficients of interest. The next model uses the indicators of $x$ and specifies latent variables ($\xi_{t}$, `xi` in the code) to represent the valid cross-sectional variance. The dependent variable in the model is one of the impercisely measured indicators of $y$.

@@

Finally, to see the benefits of removing measurement error from the dependent variable in terms of standard errors and $R^{2}$ statistics, we can specify a model with latent variables representing the valid cross-sectional variance in $y$ (`n` for $\eta$ in the code).

@@

Here, the effect $\hat{\beta}_{\eta_{t},\xi_{t}}$ is somewhat further off of the true effect of 0.3 than the preceding models. This will depend on how the latent variables are estimated, which themselves will depend on the underlying correlations between the indicators. Again, if the main goal of the model is to avoid bias, it may be advisable to just leave the manifest dependent variable as it is, and worry about measurement error in the independent variables.

@@

It is important, however, to realize that the unbiasedness of $\hat{\gamma}$ in this model is dependent on the assumption that $\E[\zeta | \bm{\xi_{t}}, \alpha_{2}] = 0$. In other words, the idiosyncratic error is mean independent of $\bm{\xi_{t}} = (\xi_{1}, \xi_{2}, ..., \xi_{T})$ as well as $\alpha_{2}$. The first part is easier to accept because we are controlling for all potential time-invariant confounders that could induce a relationship between the independent variable and the error. The unbiasedness of $\hat{\gamma}$, on the other hand rests on the assumption that the time-invariant predictor is independent of the error. If $\alpha_{2}$ represented the respondent's intelligence and $\eta_{t}$, the dependent variable, represented the respondent's income, for example, then $\hat{\gamma}$ would be biased if both were dependent on a third time-invariant variable, say level of schooling, if it is not controlled for. For this reason, we need to treat the regression on a time-invariant predictor like any other regular multivariate regression model and look to include all plausible potential confounders as controls in the model, or turn to other methods, e.g., instrumental variables.
